---
title: "STA258: Statistics with Applied Probability"
author: "Nishan Mudalige"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output:
  bookdown::html_document2:
    toc: false
    number_sections: false
knit: bookdown::render_book
documentclass: book
github-repo: nishanmudalige/STA258
bibliography: [references.bib, packages.bib]
biblio-style: apalike
link-citations: yes
nocite: '@*'
description: "Interactive landing page for STA258: Statistics with Applied Probability."
---
---

<style>
/* Landing-page layout: hide the GitBook-generated sidebar on index.html. */
.book .book-summary {
  display: none;
}

/* Re-centre the page body after the sidebar is hidden. */
.book.with-summary .book-body,
.book .book-body,
.book.with-summary .book-header.fixed,
.book .book-header.fixed {
  left: 0;
}

/* Hide the sidebar toggle button on the landing page. */
.book .book-header .btn.pull-left {
  display: none;
}

/* Keep the landing page content comfortably centred. */
.book .book-body .page-wrapper .page-inner {
  max-width: 980px;
}
</style>

# STA258: Statistics with Applied Probability <br> University of Toronto Mississauga {.unnumbered .unlisted}

```{r setup, include=FALSE}
library(rgl)
rgl::setupKnitr(autoprint = FALSE)

knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  webgl   = FALSE,
  rgl     = FALSE
)
```

<center>
```{r, echo=FALSE, warning=FALSE, webgl=TRUE, fig.show="hide"}
library(rgl); library(png); library(RCurl)

# 1. load & flip
myurl <- "https://raw.githubusercontent.com/nishanmudalige/STA258/main/Hex%20Badges/STA258_no_bg.png"
img   <- readPNG(getURLContent(myurl))
img   <- img[nrow(img):1, , ]

# 2. flatten over white
alpha      <- img[,,4]
img[,,1]   <- img[,,1] * alpha + (1 - alpha)
img[,,2]   <- img[,,2] * alpha + (1 - alpha)
img[,,3]   <- img[,,3] * alpha + (1 - alpha)
img[,,4]   <- 1

# 3. build quad
h  <- dim(img)[1]; w <- dim(img)[2]; sf <- max(w, h)/2
front <- matrix(c(
  -w/2, -h/2, 0,
  w/2, -h/2, 0,
  w/2,  h/2, 0,
  -w/2,  h/2, 0
), byrow=TRUE, ncol=3) / sf
front <- rotate3d(front, angle=pi/2, x=1, y=0, z=0)

# 4. texture coords
tex <- matrix(c(0,0, 1,0, 1,1, 0,1), byrow=TRUE, ncol=2)

# 5. draw & embed
invisible(open3d())

bg3d("white")
view3d(theta=0, phi=-90, zoom=0.675)
invisible(quads3d(
  front,
  texture   = img,
  texcoords = tex,
  lit       = FALSE,
  col       = "white"
))
```
</center>

```{r cleanup-rgl, include=FALSE}
rgl.close()
```

<h2 style="text-align: center; color: #FF2C21;">Work in Progress</h2>

---

### Nishan Mudalige, Masoud Ataei, Nurlana Alili, Xi (Bryan) Su {.unlisted .unnumbered}

This is a customized e-book for the course **STA258: Statistics with Applied Probability** at the University of Toronto Mississauga.

The site is currently being developed as an interactive learning resource for probability, statistical inference, applied modelling, and computational statistics.

<p style="text-align: center; margin-top: 1.5em;">
  <strong>STA258 Interactive Course E-book</strong><br>
  University of Toronto Mississauga
</p>

---

<!-- Script for GitHub buttons to work -->
<script async defer src="https://buttons.github.io/buttons.js"></script>

<p style="text-align: center;">
<!-- Star -->
<a class="github-button" href="https://github.com/nishanmudalige/STA258" data-color-scheme="no-preference: light; light: light; dark: dark;" data-icon="octicon-star" data-size="large" aria-label="Star nishanmudalige/STA258 on GitHub">Star</a>
<!-- Watch -->
<a class="github-button" href="https://github.com/nishanmudalige/STA258/subscription" data-color-scheme="no-preference: dark; light: light; dark: dark;" data-icon="octicon-eye" data-size="large" aria-label="Watch nishanmudalige/STA258 on GitHub">Watch</a>
<!-- Fork -->
<a class="github-button" href="https://github.com/nishanmudalige/STA258/fork" data-color-scheme="no-preference: light; light: light; dark: dark;" data-icon="octicon-repo-forked" data-size="large" aria-label="Fork nishanmudalige/STA258 on GitHub">Fork</a>
<!-- Issues -->
<a class="github-button" href="https://github.com/nishanmudalige/STA258/issues" data-color-scheme="no-preference: light; light: light; dark: dark;" data-icon="octicon-issue-opened" data-size="large" aria-label="Issue nishanmudalige/STA258 on GitHub">Issue</a>
<!-- Discuss -->
<a class="github-button" href="https://github.com/nishanmudalige/STA258/discussions" data-color-scheme="no-preference: light; light: light; dark: dark;" data-icon="octicon-comment-discussion" data-size="large" aria-label="Discuss nishanmudalige/STA258 on GitHub">Discuss</a>
</p>
