# Two Sample Hypothesis Tests

## Comparing Means with Independent Samples

<figure>

<figcaption>Structure of Two-Sample Hypothesis Tests</figcaption>
</figure>

### Setting Up Hypotheses {#setting-up-hypotheses .unnumbered}

Let $\theta_1$ and $\theta_2$ be the parameters of interest from
populations 1 and 2, respectively.

**1. Interested in whether $\theta_1 > \theta_2$:** $$\begin{aligned}
H_0\!:~ & \theta_1 = \theta_2 \\
H_a\!:~ & \theta_1 > \theta_2 \\
\text{Equivalent:} \quad & H_0\!:~ \theta_1 - \theta_2 = 0 \qquad H_a\!:~ \theta_1 - \theta_2 > 0
\end{aligned}$$

**2. Interested in whether $\theta_1 < \theta_2$:** $$\begin{aligned}
H_0\!:~ & \theta_1 = \theta_2 \\
H_a\!:~ & \theta_1 < \theta_2 \\
\text{Equivalent:} \quad & H_0\!:~ \theta_1 - \theta_2 = 0 \qquad H_a\!:~ \theta_1 - \theta_2 < 0
\end{aligned}$$

**3. Interested in whether $\theta_1 \ne \theta_2$:** $$\begin{aligned}
H_0\!:~ & \theta_1 = \theta_2 \\
H_a\!:~ & \theta_1 \ne \theta_2 \\
\text{Equivalent:} \quad & H_0\!:~ \theta_1 - \theta_2 = 0 \qquad H_a\!:~ \theta_1 - \theta_2 \ne 0
\end{aligned}$$

### Structure of a Test Statistic {#structure-of-a-test-statistic .unnumbered}

The general structure of a test statistic is:
$$\text{test statistic} = \frac{\text{(observed statistic)} - \text{(hypothesized value)}}{\text{standard error}}$$

**Common Cases:**

- $\sigma$ known: $z = \dfrac{\bar{x} - \mu_0}{\sigma / \sqrt{n}}$

- $\sigma$ unknown: $t = \dfrac{\bar{x} - \mu_0}{s / \sqrt{n}}$

- Proportions: $z = \dfrac{\hat{p} - p_0}{\sqrt{p_0(1 - p_0)/n}}$

## Hypothesis Test on a Difference of Means ($\mu_1 - \mu_2$) {#hypothesis-test-on-a-difference-of-means-mu_1---mu_2 .unnumbered}

- When $\sigma_1$ and $\sigma_2$ are known:

$$\begin{aligned}
H_0&: \mu_1 - \mu_2 = 0 \\
H_a&: \mu_1 - \mu_2 > 0 \quad \text{(or } < 0 \text{, } \ne 0 \text{)}
\end{aligned}$$

**Test Statistic:**
$$z = \frac{(\bar{x}_1 - \bar{x}_2) - 0}{\sqrt{\frac{\sigma_1^2}{n_1} + \frac{\sigma_2^2}{n_2}}}$$

Reference distribution: standard normal (Z distribution).

*Note: This scenario is rare in practice since both population standard
deviations $\sigma_1$ and $\sigma_2$ are seldom known.*

- When $\sigma_1$ and $\sigma_2$ are unknown:

**Case 1: Assume equal variances $\sigma_1^2 = \sigma_2^2$**
$$\begin{aligned}
H_0&: \mu_1 - \mu_2 = 0 \\
H_a&: \mu_1 - \mu_2 > 0 \quad \text{(or } < 0 \text{, } \ne 0 \text{)}
\end{aligned}$$

**Test Statistic:**
$$t = \frac{(\bar{x}_1 - \bar{x}_2) - 0}{s_p \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}}$$

where the pooled standard deviation is defined as:
$$s_p^2 = \frac{(n_1 - 1)s_1^2 + (n_2 - 1)s_2^2}{n_1 + n_2 - 2}$$

Reference distribution: $t$-distribution with $n_1 + n_2 - 2$ degrees of
freedom.

(Use this only when equal variances can reasonably be assumed.)\

**Case 2: Assume Unequal Variances $\sigma_1^2 \ne \sigma_2^2$ (Welch's
t-test)** $$\begin{aligned}
H_0 &: \mu_1 - \mu_2 = 0 \\
H_a &: \mu_1 - \mu_2 > 0 \quad (\text{or } < 0,\ \ne 0)
\end{aligned}$$

**Test Statistic:**
$$t^* = \frac{(\bar{x}_1 - \bar{x}_2) - 0}{\sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}}$$

Reference distribution: approximately a $t$-distribution.

*Degrees of freedom (by hand):* $$\min(n_1 - 1,\ n_2 - 1)$$

*Note:* R uses a more sophisticated approximation for degrees of freedom
in this case.

## Comparing Means of Independent Samples (Normal Population Assumptions) {#comparing-means-of-independent-samples-normal-population-assumptions .unnumbered}

We should check the assumption that the underlying populations of
individual responses are each Normally distributed. Nearly Normal
Condition:

- We must check this for both groups; a violation for either one
  violates the condition.

- The Normality assumption matters most when sample sizes are very
  small.

- For $n < 10$ in either group, this method should not be used if the
  histogram or Normality plots show clear skewness.

- For $n$'s of 10 or so, a moderately skewed histogram is okay. But, for
  strongly skewed data or data containing outliers this method should be
  avoided.

- For larger samples $n \geq 20$, data skewness is less of an issue ---
  but, we still need to check if there are any outliers in the data,
  extreme skewness, and multiple modes.

### Comparing Two Populations Means: Independent Sampling (Equal Variances Assumed)

Consider two independent populations with unknown means $\mu_1$ and
$\mu_2$, and unknown standard deviations $\sigma_1$ and $\sigma_2$
($\sigma_1 = \sigma_2$), respectively. We can make an inference about
their mean difference $\mu_1 - \mu_2$ by using the difference between
their point estimates (sample means): $\bar{Y}_1 - \bar{Y}_2$. When the
assumptions and conditions are met,

$$\frac{(\bar{Y}_1 - \bar{Y}_2) - (\mu_1 - \mu_2)}{S_p \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}}
= \frac{(\bar{Y}_1 - \bar{Y}_2) - (\mu_1 - \mu_2)}{\sqrt{S_p^2 \left[\frac{1}{n_1} + \frac{1}{n_2} \right] }},$$

can be modelled by a $t(\nu)$ distribution; where $\nu = n_1 + n_2 - 2$
and

$$S_p^2 = \frac{(n_1 - 1)s_1^2 + (n_2 - 1)s_2^2}{n_1 + n_2 - 2}.$$

::: tcolorbox
**Conditions Required for Valid Inference about $\mu_1 - \mu_2$:**

1.  The two samples are randomly selected in an independent manner from
    the two target populations.

2.  Both sampled populations have distributions that are approximately
    Normal.

3.  The population variances are equal (e.g., $\sigma_1 = \sigma_2$).
:::

### Small-Sample Confidence Interval for $\mu_1 - \mu_2$ (with equal variances) {#small-sample-confidence-interval-for-mu_1---mu_2-with-equal-variances .unnumbered}

**Parameter:** $\mu_1 - \mu_2$

**Confidence interval** ($\nu = \text{df}$):

$$(\bar{Y}_1 - \bar{Y}_2) \pm t_{\alpha/2}(\nu) S_p \sqrt{\frac{1}{n_1} + \frac{1}{n_2}},$$

where $\nu = n_1 + n_2 - 2$ and
$$S_p^2 = \frac{(n_1 - 1)s_1^2 + (n_2 - 1)s_2^2}{n_1 + n_2 - 2}$$

(requires that Normal samples are independent and the assumption that
$\sigma_1^2 = \sigma_2^2$). The critical value $t_{\alpha/2}(\nu)$
depends on the particular confidence level and the number of degrees of
freedom.

:::::::::::::::::::: example
Comparing Two Population Means Managerial Success Indexes for Two Groups
(With Equal Variances Assumed)

Behavioural researchers have developed an index designed to measure
managerial success. The index (measured on a 100-point scale) is based
on the manager's length of time in the organization and their level
within the term; the higher the index, the more successful the manager.
Suppose a researcher wants to compare the average index for the two
groups of managers at a large manufacturing plant. Managers in group 1
engage in **high volume of interactions** with people outside the
managers' work unit (such interaction include phone and face-to-face
meetings with customers and suppliers, outside meetings, and public
relation work). Managers in group 2 **rarely interact** with people
outside their work unit. Independent random samples of 12 and 15
managers are selected from groups 1 and 2, respectively, and success
index of each is recorded.

**Response variable:** Managerial Success Indexes (quantitative,
continuous, 0--100 scale)

**Explanatory variable:** Type of group (nominal categorical: *Group 1 =
interaction with outsiders*, *Group 2 = fewer interactions*)

**R Code**

::: tcolorbox
    # Importing data file into R
    success = read.csv(file = "success.csv", header = TRUE);

    # Getting names of variables
    names(success);

    # Seeing first few observations
    head(success);

    # Attaching data file
    attach(success);
:::

**R Code**

::: tcolorbox
    ## [1] "Success_Index" "Group"
    ##   Success_Index Group
    ## 1            65     1
    ## 2            66     1
    ## 3            58     1
    ## 4            70     1
    ## 5            78     1
    ## 6            53     1
:::

**R code (Descriptive Statistics)**

::: tcolorbox
    # loading library mosaic
    library(mosaic)

    favstats(Success_Index ~ Group)
:::

**R Code (Descriptive Statistics)**

::: tcolorbox
    ## .group min   Q1 median   Q3 max     mean       sd  n
    ##       1  53 62.25  65.50 69.25  78 65.33333 6.610368 12
    ##       2  34 42.50  50.00 54.50  68 49.46667 9.334014 15
:::

**R Code (Descriptive Statistics)**

::: tcolorbox
    summary(Success_Index[Group == 1]);
    length(Success_Index[Group == 1]);
    sd(Success_Index[Group == 1]);

    summary(Success_Index[Group == 2]);
    length(Success_Index[Group == 2]);
    sd(Success_Index[Group == 2]);
:::

Note: Group 1 = "interaction with outsiders" and Group 2 = "fewer
interactions".

**R Output**

::: tcolorbox
    ##  Min.  1st Qu.  Median    Mean  3rd Qu.    Max. 
    ##  53.00   62.25   65.50   65.33   69.25   78.00 
    ## [1] 12
    ## [1] 6.610368

    ##  Min.  1st Qu.  Median    Mean  3rd Qu.    Max. 
    ##  34.00   42.50   50.00   49.47   54.50   68.00 
    ## [1] 15
    ## [1] 9.334014
:::

**Nearly Normal Condition (Group 1: "interaction with outsiders"):**

::: tcolorbox
    stem(Success_Index[Group == 1]);
:::

**R Output**

::: tcolorbox
    ## 
    ##  The decimal point is 1 digit(s) to the right of the |
    ## 
    ##  5 | 38
    ##  6 | 0335689
    ##  7 | 018
:::

**Nearly Normal Condition (Group 2: "fewer interactions"):**

::: tcolorbox
    stem(Success_Index[Group == 2]);
:::

**R Output**

::: tcolorbox
    ## 
    ##  The decimal point is 1 digit(s) to the right of the |
    ## 
    ##  3 | 46
    ##  4 | 22368
    ##  5 | 023367
    ##  6 | 28
:::

**Nearly Normal Condition (Group 1: "interaction with outsiders"):**

::: tcolorbox
    qqnorm(Success_Index[Group == 1]);
    qqline(Success_Index[Group == 1]);
:::
```{r fig.cap="Q-Q Plot for Group 1: Interaction with Outsiders", echo=FALSE, fig.align='center', out.width='60%'}
# Recreate Group 1 data manually
success_group1 <- c(65, 66, 58, 70, 78, 53, 62, 64, 68, 69, 60, 67)  # 12 values


qqnorm(success_group1)
qqline(success_group1)
```

**Nearly Normal Condition (Group 2: "fewer interactions"):**

::: tcolorbox
    qqnorm(Success_Index[Group == 2]);
    qqline(Success_Index[Group == 2]);
:::

```{r fig.cap="Q-Q Plot for Group 2: Fewer Interactions", echo=FALSE, fig.align='center', out.width='60%'}
# Step 1: Recreate Group 2 data manually
success_group2 <- c(34, 40, 45, 46, 50, 42, 44, 48, 54, 49, 52, 53, 55, 60, 68)  # 15 values


qqnorm(success_group2)
qqline(success_group2)
```

:::

**Nearly Normal Condition:**

::: tcolorbox
    boxplot(Success_Index ~ Group, col = c("red", "blue"))
:::
```{r fig.cap="Boxplot of Success Index by Group", echo=FALSE, fig.align='center', out.width='60%'}
# Create Group 1 and Group 2 manually
group1 <- c(65, 66, 58, 70, 78, 53, 62, 64, 68, 69, 60, 67)
group2 <- c(34, 40, 45, 46, 50, 42, 44, 48, 54, 49, 52, 53, 55, 60, 68)

# Combine into full dataset
Success_Index <- c(group1, group2)
Group <- c(rep(1, length(group1)), rep(2, length(group2)))


boxplot(Success_Index ~ Group, col = c("red", "blue"))
```


[]{#fig:boxplot-success label="fig:boxplot-success"}
:::

**Boxplot with `ggplot2`:**

::: tcolorbox
    # loading library;
    library(ggplot2);

    # converting a numeric variable into factor (categorical data)
    group <- factor(Group);

    # bp: just a name (not code) to store boxplots;
    bp <- ggplot(success,
                 aes(x = group, y = Success_Index, fill = group));

    our.labs <- c("Interaction with Outsiders", "Fewer Interactions");

    bp +
      geom_boxplot() +
      scale_x_discrete(labels = our.labs);
:::
```{r fig.cap="Boxplot of Success Index by Group using ggplot2", echo=FALSE, fig.align='center', out.width='60%'}
# Load required library
library(ggplot2)

# Step 1: Manually define data
group1 <- c(65, 66, 58, 70, 78, 53, 62, 64, 68, 69, 60, 67)
group2 <- c(34, 40, 45, 46, 50, 42, 44, 48, 54, 49, 52, 53, 55, 60, 68)
Success_Index <- c(group1, group2)
Group <- c(rep(1, length(group1)), rep(2, length(group2)))

# Step 2: Convert to data frame and factor
success <- data.frame(Success_Index, Group)
success$Group <- factor(success$Group)

# Step 3: Create and print plot
ggplot(success, aes(x = Group, y = Success_Index, fill = Group)) +
  geom_boxplot() +
  scale_x_discrete(labels = c("Interaction with Outsiders", "Fewer Interactions")) +
  labs(x = "Group", y = "Success Index") +
  theme_minimal()
```


**Checking the Assumptions and Conditions**

**Independent Group Assumption:** The success index in group 1 is
unrelated to the success index in group 2.

**Randomization Condition:** The 27 managers were randomly and
independently selected (12 for group 1, and 15 for group 2).

**Nearly Normal Condition:** The two boxplots of success indexes do not
show skewness; the two stemplots/histograms of success indexes are
unimodal, fairly symmetric and approximately bell-shaped. Q--Q plots
also suggest the normality assumption is reasonable.

**Equal Variances Assumption:** The two boxplots of success indexes
appear to have the same spread; thus, the samples appear to have come
from populations with approximately the same variance.

Since the conditions are satisfied, it is appropriate to construct a $t$
confidence interval with\
$df = 12 + 15 - 2 = 25$.\
**From the data, the following statistics were calculated:**

$$\begin{aligned}
n_1 &= 12 &\quad n_2 &= 15 \\
\bar{x}_1 &= 65.33 &\quad \bar{x}_2 &= 49.47 \\
s_1^2 &= 6.61^2 &\quad s_2^2 &= 9.33^2
\end{aligned}$$

**The pooled variance estimator is:**

$$s_p^2 = \frac{(n_1 - 1)s_1^2 + (n_2 - 1)s_2^2}{n_1 + n_2 - 2} 
= \frac{(12 - 1)(6.61^2) + (15 - 1)(9.33^2)}{12 + 15 - 2}
= 67.97$$

**The number of degrees of freedom is:**
$$\nu = n_1 + n_2 - 2 = 12 + 15 - 2 = 25$$

**Two-sample t-test (Student's t-test) for the Difference Between Means
$\mu_1 - \mu_2$:**

Is there evidence to suggest that the mean success index differs between
the two groups?

**1. State Hypotheses:** $$\begin{aligned}
H_0 &: \mu_1 = \mu_2 \quad \text{or} \quad H_0 : \mu_1 - \mu_2 = 0 \\
H_a &: \mu_1 \ne \mu_2 \quad \text{or} \quad H_a : \mu_1 - \mu_2 \ne 0
\end{aligned}$$

**2. Test Statistic:**
$$t^* = \frac{\bar{x}_1 - \bar{x}_2}{\sqrt{\frac{s_p^2}{n_1} + \frac{s_p^2}{n_2}}}
= \frac{65.33 - 49.47}{\sqrt{\frac{67.97}{12} + \frac{67.97}{15}}}
= 4.97$$

**3. P-value:**

Using table:
$$\text{df} = 25 \quad \Rightarrow \quad \text{P-value} < 2(0.005)$$

Using R:

::: tcolorbox
    # One way:
    2*(1 - pt(4.97, df = 25));

    # Another way:
    2 * pt(4.97, df = 25, lower.tail = FALSE);
:::

Both methods return: $$\text{P-value} \approx 4.03 \times 10^{-5}$$ **4.
Conclusion.**

Since the P-value is very small, we have strong evidence to indicate
that there is a difference in mean success index between group 1 and
group 2.

**Note.** As a follow-up, we could find a 95% CI for $\mu_1 - \mu_2$ to
estimate this difference. Then, we could provide an estimate of how much
higher the mean success index for group 1 is.
::::::::::::::::::::

### Comparing Two Populations Means: Independent Sampling (Unequal Variances Assumed)

### Small-Sample Confidence Interval for $\mu_1 - \mu_2$ (Unequal Variances) {#small-sample-confidence-interval-for-mu_1---mu_2-unequal-variances .unnumbered}

Draw an SRS of size $n_1$ from a Normal population with unknown mean
$\mu_1$, and draw an independent SRS of size $n_2$ from another Normal
population with unknown mean $\mu_2$. A confidence interval for
$\mu_1 - \mu_2$ is given by

$$(\bar{x}_1 - \bar{x}_2) \pm t^* \sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}$$

Here $t^*$ is the critical value for the $t(k)$ density curve with area
$C$ between $-t^*$ and $t^*$. The degrees of freedom $k$ are equal to
the smaller of $n_1 - 1$ and $n_2 - 1$.

### Degrees of Freedom {#degrees-of-freedom .unnumbered}

*Option 1.* With software, use the statistic $t$ with accurate critical
values from the approximating $t$ distribution.

The distribution of the two-sample $t$ statistic is very close to the
$t$ distribution with degrees of freedom $df$ given by

$$df = \frac{\left( \frac{s_1^2}{n_1} + \frac{s_2^2}{n_2} \right)^2}
{
\left( \frac{1}{n_1 - 1} \left( \frac{s_1^2}{n_1} \right)^2 \right)
+
\left( \frac{1}{n_2 - 1} \left( \frac{s_2^2}{n_2} \right)^2 \right)
}$$

This approximation is accurate when both sample sizes $n_1$ and $n_2$
are 5 or larger.

*Option 2.* Without software, use the statistic $t$ with critical values
from the $t$ distribution with degrees of freedom equal to the smaller
of $n_1 - 1$ and $n_2 - 1$. These procedures are always conservative for
any two Normal populations.

### Two-Sample t-Test (Unequal Variances) {#two-sample-t-test-unequal-variances .unnumbered}

To test the hypothesis $H_0 : \mu_1 = \mu_2$, calculate the two-sample
$t$ statistic:

$$t^* = \frac{\bar{x}_1 - \bar{x}_2}{\sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}}$$

and use $P$-values or critical values for the $t(k)$ distribution.

::: example
"Conservationists have despaired over destruction of tropical rain
forest by logging, clearing, and burning". These words begin a report on
a statistical study of the effects of logging in Borneo. Here are data
on the number of tree species in 12 unlogged forest plots and 9 similar
plots logged 8 years earlier:

Unlogged: 22 18 22 20 15 21 13 13 19 13 19 15\
Logged : 17 4 18 14 18 15 15 10 12

Does logging significantly reduce the mean number of species in a plot
after 8 years? State the hypotheses and do a t test. Is the result
significant at the 5% level?\
**Solution:**\
1. State hypotheses. $H_0: \mu_1 = \mu_2$ vs. $H_a: \mu_1 > \mu_2$,
where $\mu_1$ is the mean number of species in unlogged plots and
$\mu_2$ is the mean number of species in plots logged 8 years earlier.

2\. Test statistic.
$$t^* \;=\;\frac{\bar x_1 - \bar x_2}{\sqrt{s_1^2/n_1 + s_2^2/n_2}}
\;=\;2.1140$$
$$(\bar x_1 = 17.5,\;\bar x_2 = 13.6666,\;s_1 = 3.5290,\;s_2 = 4.5,\;n_1 = 12,\;n_2 = 9)$$

3\. P-value. Using Table, we have $df = 8$, and
$0.025 < \text{P-value} < 0.05$.

4\. Conclusion. Since P-value $< 0.05$, we reject $H_0$. There is strong
evidence that the mean number of species in unlogged plots is greater
than that for logged plots 8 years after logging.
:::

:::::::::::::::: example
A company that sells educational materials reports statistical studies
to convince customers that its materials improve learning. One new
product supplies "directed reading activities" for classroom use. These
activities should improve the reading ability of elementary school
pupils.

A consultant arranges for a third-grade class of 21 students to take
part in these activities for an eight-week period. A control classroom
of 23 third-graders follows the same curriculum without the activities.
At the end of the eight weeks, all students are given a Degree of
Reading Power (DRP) test, which measures the aspects of reading ability
that the treatment is designed to improve. The data appear in the
following table.

::: center
+---------------------------+-------------------------------+
| **Treatment**             | **Control**                   |
+:====:+:====:+:====:+:====:+:=====:+:=====:+:=====:+:=====:+
| 24   | 61   | 59   | 46   | 42    | 33    | 46    | 37    |
+------+------+------+------+-------+-------+-------+-------+
| 43   | 44   | 52   | 43   | 43    | 41    | 10    | 42    |
+------+------+------+------+-------+-------+-------+-------+
| 58   | 67   | 62   | 57   | 55    | 19    | 17    | 55    |
+------+------+------+------+-------+-------+-------+-------+
| 71   | 49   | 54   |      | 26    | 54    | 60    | 28    |
+------+------+------+------+-------+-------+-------+-------+
| 43   | 53   | 57   |      | 62    | 20    | 53    | 48    |
+------+------+------+------+-------+-------+-------+-------+
| 49   | 56   | 33   |      | 37    | 85    | 42    |       |
+------+------+------+------+-------+-------+-------+-------+
:::

Because we hope to show that the treatment (Group 1) is better than the
control (Group 2), the hypotheses are:

$$H_0 : \mu_1 = \mu_2$$ $$H_a : \mu_1 > \mu_2$$

::: tcolorbox
    # Step 1. Entering data
    treatment = c(24, 61, 59, 46, 43, 44, 52, 43, 58, 67, 62, 57, 
                  71, 49, 54, 43, 53, 57, 49, 56, 33);
    control = c(42, 33, 46, 37, 43, 41, 10, 42, 55, 19, 17, 55, 
                26, 54, 60, 28, 62, 20, 53, 48, 37, 85, 42);
:::

**Nearly Normal Condition (treatment):**

::: tcolorbox
    # Making stemplot;

    stem(treatment);
:::

::: tcolorbox
    ##
    ## The decimal point is 1 digit(s) to the right of the |
    ##
    ## 2 | 4
    ## 3 | 3
    ## 4 | 3334699
    ## 5 | 23467789
    ## 6 | 127
    ## 7 | 1
:::

**Nearly Normal Condition (control):**

::: tcolorbox
    # Making stemplot;

    stem(control);
:::

::: tcolorbox
    ##
    ## The decimal point is 1 digit(s) to the right of the |
    ##
    ## 0 | 079
    ## 2 | 068377
    ## 4 | 12223683455
    ## 6 | 02
    ## 8 | 5
:::

**Nearly Normal Condition (treatment):**

::: tcolorbox
    # Making Q-Q plot;
    qqnorm(treatment, pch=19, col="red", main="Treatment");
    qqline(treatment, lty=2);
:::
```{r fig.cap="Q-Q Plot for Treatment Group", echo=FALSE, fig.align='center', out.width='60%'}
# Step 1: Enter data
treatment <- c(24, 61, 59, 46, 43, 44, 52, 43, 58, 67, 62, 57,
               71, 49, 54, 43, 53, 57, 49, 56, 33)

# Step 2: Generate Q-Q plot
qqnorm(treatment, pch = 19, col = "red", main = "")
qqline(treatment, lty = 2, col = "blue")
```


**Nearly Normal Condition (control):**

::: tcolorbox
    # Making Q-Q plot;
    qqnorm(treatment, pch=19, col="red", main="Control");
    qqline(control, lty=2);
:::

```{r fig.cap="Q-Q Plot for Control Group", echo=FALSE, fig.align='center', out.width='60%'}
# Step 1: Enter data
control <- c(42, 33, 46, 37, 43, 41, 10, 42, 55, 19, 17, 55,
             26, 54, 60, 28, 62, 20, 53, 48, 37, 85, 42)
# Step 2: Generate Q-Q plot
qqnorm(control, pch = 19, col = "red", main = "")
qqline(control, lty = 2, col = "blue")
```

:::

Stemplots suggest that there is a mild outlier in the control group but
no deviation from Normality serious enough to prevent us from using t
procedures. Normal Q-Q plots for both groups confirm that both are
roughly Normal. The summary statistics are:

::: tcolorbox
    summary(treatment);

    ## Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
    ## 24.00   44.00   53.00   51.48   58.00   71.00

    summary(control);

    ## Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
    ## 10.00   30.50   42.00   41.52   53.50   85.00
:::

::: tcolorbox
    # Step 1. Entering data;

    treatment = c(24, 61, 59, 46, 43, 44, 52, 43, 58, 67, 62, 57, 
                  71, 49, 54, 43, 53, 57, 49, 56, 33);
                  
    control = c(42, 33, 46, 37, 43, 41, 10, 42, 55, 19, 17, 55, 
                26, 54, 60, 28, 62, 20, 53, 48, 37, 85, 42);

    # Step 2. Hypothesis Test

    t.test(treatment, control, alternative="greater")
:::

**HT (using R)**

::: tcolorbox
    ##
    ##  Welch Two Sample t-test
    ## 
    ## data:  treatment and control
    ## t = 2.3106, df = 37.855, p-value = 0.01305
    ## alternative hypothesis: true difference in means is greater than 0
    ## 95 percent confidence interval:
    ##  2.333784      Inf
    ## sample estimates:
    ## mean of x mean of y 
    ##    51.47619 41.52174 
:::

**HT (using table)**

::: tcolorbox
    round (mean(treatment) ,2);

    ##[1] 51.48

    round (sd(treatment) ,2);

    ##[1] 11.01

    round (mean(control) ,2);

    ##[1] 41.52

    round (sd(control) ,2);

    ##[1] 17.15
:::

Test statistic.\
$$t^* = \frac{\bar{x}_1 - \bar{x}_2}{\sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}} = 2.31$$
$$\quad
(\bar{x}_1 = 51.48, \, \bar{x}_2 = 41.52, \, s_1 = 11.01, \, s_2 = 17.15,
\quad n_1 = 21 \text{ and } n_2 = 23)$$

The conservative approach uses the *t*(20) distribution. The P-value for
the one-sided test is $$\text{P-value} = P\bigl(T \geq 2.31 \bigr)$$
Comparing $t = 2.31$ with the entries in Table 5 for 20 degrees of
freedom, we see that $$0.01 < \text{P-value} < 0.025.$$

Since our P-value is "small", we reject the null hypothesis (Note that
we would reject $H_0$ at the 2.5% significance level). The data strongly
suggest that directed reading activity improves the DRP score.

The design of the DRP study is not ideal. Random assignment of students
was not possible in a school environment, so existing third-grade
classes were used. The effect of the reading programs is therefore
confounded with any other differences between the two classes. The
classes were chosen to be as similar as possible in variables such as
the social and economic status of the students. Pretesting showed that
the two classes were on the average quite similar in reading ability at
the beginning of the experiment. To avoid the effect of two different
teachers, the same teacher taught reading in both classes during the
eight-week period of the experiment. We can therefore be somewhat
confident that our two-sample procedure is detecting the effect of the
treatment and not some other difference between the classes.
::::::::::::::::

::: nt
**Why do we use hypothesis tests?**

To conduct tests on parameters and to quantify the degree of certainty
with probability.

The smaller the p-value, the stronger the evidence against $H_0$
(provided assumptions are met).
:::

**Understanding Significance Levels**

Suppose you have the following limited information:

- A test was statistically significant at the 5% level.

We **cannot** make this conclusion for certain. That is,
$$p\text{-value} < 0.05 \quad \not\Rightarrow \quad p\text{-value} < 0.01.$$

- A test was statistically significant at the 1% level.

Yes! Because $$p\text{-value} < 0.01 < 0.05.$$

## The Fold Rule

A rule which can be used to quickly determine whether population
variances are equal or unequal using sample variances.

*Note: only a rule*

If $$\frac{\max(s_1, s_2)}{\min(s_1, s_2)} < \sqrt{2} 
\quad \text{then we can consider } \sigma_1^2 = \sigma_2^2$$

and

$$\frac{\max(s_1^2, s_2^2)}{\min(s_1^2, s_2^2)} < 2 
\quad \text{then we can consider } \sigma_1^2 = \sigma_2^2$$

**Note:** rather crude technique.

Hypothesis tests for equality of two variances do exist:

$$H_0: \sigma_1^2 = \sigma_2^2 \quad \text{vs} \quad H_a: \sigma_1^2 \neq \sigma_2^2$$

These tests involve a $\chi^2$ test statistic.

## Two Sample Hypothesis Test on Paired Data

When observations in sample 1 matches with an observation in sample 2.
Observations in sample 1 are, usually, highly, correlated with
observations in sample 2, these data are often called matched pairs. For
each pair (the same cases), we form: Difference = observation in sample
2 - observation in sample 1. Thus, we have one single sample of
differences scores. For example, in longitudinal studies: Pre- and
post-survey of attitudes towards statistics (Same student is measured
twice: Time 1 (pre) and Time 2 (post). We measure change in the
attitudes: Post - Pre (for each student). Often these types of studies
are called, repeated measures.\
Paired Data Condition: the data must be quantitative and paired.\
Independence Assumption:

- If the data are p aired, the groups are not independent. For this
  methods, it is the differences that must be independent of each other.

- The pairs may be a random sample.

- In experimental design, the order of the two treatments may be
  randomly assigned, or the treatments ma y be randomly assigned to one
  member of each pair.

- In a before-and-after study, we may believe that the observed
  differences are representative sample of a population of interest. If
  there is any doubt, we need to include a control group to be able to
  draw conclusions.

- If samples are bigger than 10 % of the target population, we need to
  acknowledge this and note in our report. When we sample from a ﬁnite
  population, we should be careful not to sample more than 10 % of that
  population. Sampling too large a fraction of the population calls the
  independence assumption into question.

Recall chapter 10 when we first introduced paired data, now we are going
to use it again:

::: center
<figure>
<table>
<tbody>
<tr>
<td style="text-align: center;">Sample Units</td>
<td style="text-align: center;">Measurement 1 (<span
class="math inline"><em>M</em><sub>1</sub></span>)</td>
<td style="text-align: center;">Measurement 2 (<span
class="math inline"><em>M</em><sub>2</sub></span>)</td>
<td style="text-align: center;">Difference (<span
class="math inline"><em>M</em><sub>2</sub> − <em>M</em><sub>1</sub></span>
or <span
class="math inline"><em>M</em><sub>1</sub> − <em>M</em><sub>2</sub></span>)</td>
</tr>
<tr>
<td style="text-align: center;">1</td>
<td style="text-align: center;"><span
class="math inline"><em>x</em><sub>11</sub></span></td>
<td style="text-align: center;"><span
class="math inline"><em>x</em><sub>12</sub></span></td>
<td style="text-align: center;"><span
class="math inline"><em>x</em><sub><em>d</em>1</sub> = <em>x</em><sub>12</sub> − <em>x</em><sub>11</sub></span></td>
</tr>
<tr>
<td style="text-align: center;">2</td>
<td style="text-align: center;"><span
class="math inline"><em>x</em><sub>21</sub></span></td>
<td style="text-align: center;"><span
class="math inline"><em>x</em><sub>22</sub></span></td>
<td style="text-align: center;"><span
class="math inline"><em>x</em><sub><em>d</em>2</sub> = <em>x</em><sub>22</sub> − <em>x</em><sub>21</sub></span></td>
</tr>
<tr>
<td style="text-align: center;">3</td>
<td style="text-align: center;"><span
class="math inline"><em>x</em><sub>31</sub></span></td>
<td style="text-align: center;"><span
class="math inline"><em>x</em><sub>32</sub></span></td>
<td style="text-align: center;"><span
class="math inline"><em>x</em><sub><em>d</em>2</sub> = <em>x</em><sub>32</sub> − <em>x</em><sub>31</sub></span></td>
</tr>
<tr>
<td style="text-align: center;">......</td>
<td style="text-align: center;"></td>
<td style="text-align: center;"></td>
<td style="text-align: center;"></td>
</tr>
<tr>
<td style="text-align: center;">n</td>
<td style="text-align: center;"><span
class="math inline"><em>x</em><sub><em>n</em>1</sub></span></td>
<td style="text-align: center;"><span
class="math inline"><em>x</em><sub><em>n</em>2</sub></span></td>
<td style="text-align: center;"><span
class="math inline"><em>x</em><sub><em>d</em><em>n</em></sub> = <em>x</em><sub><em>n</em>2</sub> − <em>x</em><sub><em>n</em>1</sub></span></td>
</tr>
</tbody>
</table>
<figcaption>A table of paired data</figcaption>
</figure>
:::

From that table, we can get $\bar{X_d}$, which is the mean, variance and
standard deviation of the difference. We need these values to continue
our analysis.\
**Step 1: Stating the Structure of Testing Hypothesis**

The idea of two sample hypothesis test on pair data is transforming it
to one sample hypothesis data, so that our analysis based on $\mu_d$.

::: center
<figure>
<table>
<tbody>
<tr>
<td style="text-align: center;">Cases</td>
<td style="text-align: center;">Null Hypothesis</td>
<td style="text-align: center;">Alternative Hypothesis</td>
</tr>
<tr>
<td style="text-align: center;">1</td>
<td style="text-align: center;"><span
class="math inline"><em>μ</em><sub><em>d</em></sub> = 0</span></td>
<td style="text-align: center;"><span
class="math inline"><em>μ</em><sub><em>d</em></sub> &gt; 0</span></td>
</tr>
<tr>
<td style="text-align: center;">2</td>
<td style="text-align: center;"><span
class="math inline"><em>μ</em><sub><em>d</em></sub> = 0</span></td>
<td style="text-align: center;"><span
class="math inline"><em>μ</em><sub><em>d</em></sub> &lt; 0</span></td>
</tr>
<tr>
<td style="text-align: center;">3</td>
<td style="text-align: center;"><span
class="math inline"><em>μ</em><sub><em>d</em></sub> = 0</span></td>
<td style="text-align: center;"><span
class="math inline"><em>μ</em><sub><em>d</em></sub> ≠ 0</span></td>
</tr>
</tbody>
</table>
<figcaption>All possible cases of two sample hypothesis test on paired
data</figcaption>
</figure>
:::

**Step 2: Computing Test Statistics**

::: definition
The test statistics of two sample hypothesis test on paired data is
given by:
$$t_* = \frac{\bar{x_d}-0}{s_d / \sqrt{n}} \sim t_{n-1;\alpha/2}.$$
$\bar{x_d}$ is the mean of sample difference and $s_d$ is the standard
deviation of difference data.
:::

**Step 3: Finding the P-value**

Two sample hypothesis test is quite similar to one sample hypothesis
test on a mean. Notes that we are working within 2 measurement. You can
refer one sample hypothesis test to get the idea about find the p-value
in this case.\
**Step 4: Comparing P-value with $\alpha$-level**

If p-value is less than $\alpha$-level, then we reject the null
hypothesis ($H_0$) and accept the alternative hypothesis ($H_a$).
Otherwise, If p-value is greater than $\alpha$-level, then we do not
reject the null hypothesis ($H_0$) and reject the alternative hypothesis
($H_a$).\
**Step 5: Final Conclusion** If we reject the null hypothesis, then we
conclude that: there is sufficient evidence to reject the null
hypothesis. If we do not reject the null hypothesis, then we conclude
that: there is insufficient evidence to reject the null hypothesis.\
Note that your final conclusion is based on the value of $\mu_d$, but we
still need more. We also can know which group has larger means from the
result ($\mu_d$).

- If $\mu_d > 0$: from the table above we get $M_2-M_1 >0$, then
  $M_2 > M_1$.

- If $\mu_d < 0$: from the table above we get $M_2-M_1 <0$, then
  $M_2 < M_1$.

- If $\mu_d \neq 0$: from the table above we get $M_2-M_1 \neq 0$, then
  $M_2 \neq M_1$.

:::: example
In an effort to determine whether a new type of fertilizer is more
effective than the type currently in use, researchers took 12 two-acre
plots of land scattered throughout the county. Each plot was divided
into two equal-size sub plots, one of which was treated with the new
fertilizer. Wheat was planted, and the crop yields were measured.

::: center
<figure>
<table>
<tbody>
<tr>
<td style="text-align: center;">Plot</td>
<td style="text-align: center;"><span class="math inline">1</span></td>
<td style="text-align: center;"><span class="math inline">2</span></td>
<td style="text-align: center;"><span class="math inline">3</span></td>
<td style="text-align: center;"><span class="math inline">4</span></td>
<td style="text-align: center;"><span class="math inline">5</span></td>
<td style="text-align: center;"><span class="math inline">6</span></td>
<td style="text-align: center;"><span class="math inline">7</span></td>
<td style="text-align: center;"><span class="math inline">8</span></td>
<td style="text-align: center;"><span class="math inline">9</span></td>
<td style="text-align: center;"><span class="math inline">10</span></td>
<td style="text-align: center;"><span class="math inline">11</span></td>
<td style="text-align: center;"><span class="math inline">12</span></td>
</tr>
<tr>
<td style="text-align: center;">Current</td>
<td style="text-align: center;"><span class="math inline">56</span></td>
<td style="text-align: center;"><span class="math inline">45</span></td>
<td style="text-align: center;"><span class="math inline">68</span></td>
<td style="text-align: center;"><span class="math inline">72</span></td>
<td style="text-align: center;"><span class="math inline">61</span></td>
<td style="text-align: center;"><span class="math inline">69</span></td>
<td style="text-align: center;"><span class="math inline">57</span></td>
<td style="text-align: center;"><span class="math inline">55</span></td>
<td style="text-align: center;"><span class="math inline">60</span></td>
<td style="text-align: center;"><span class="math inline">72</span></td>
<td style="text-align: center;"><span class="math inline">75</span></td>
<td style="text-align: center;"><span class="math inline">66</span></td>
</tr>
<tr>
<td style="text-align: center;">New</td>
<td style="text-align: center;"><span class="math inline">60</span></td>
<td style="text-align: center;"><span class="math inline">49</span></td>
<td style="text-align: center;"><span class="math inline">66</span></td>
<td style="text-align: center;"><span class="math inline">73</span></td>
<td style="text-align: center;"><span class="math inline">59</span></td>
<td style="text-align: center;"><span class="math inline">67</span></td>
<td style="text-align: center;"><span class="math inline">61</span></td>
<td style="text-align: center;"><span class="math inline">60</span></td>
<td style="text-align: center;"><span class="math inline">58</span></td>
<td style="text-align: center;"><span class="math inline">75</span></td>
<td style="text-align: center;"><span class="math inline">72</span></td>
<td style="text-align: center;"><span class="math inline">68</span></td>
</tr>
</tbody>
</table>
<figcaption>Data of example 14.1</figcaption>
</figure>
:::

Can we conclude at the 5% signiﬁcance level that the new fertilizer is
more eﬀective than the current one?

**Solution:**

You can verify that the mean and standard deviation of the twelve
difference measurements are $\bar{d} = new - current = 1$ and
$s_d = 3.0151$.

**Step 1: State Hypothesis**

$H_0: \mu_d = 0$ and $H_a: \mu_d>0$

**Step 2: Find test statistics**

$t* = \frac{\bar{d} - 0}{s_d/\sqrt{n}} = \frac{1}{3.0151/\sqrt{12}} = 1.1489$

**Step 3: Compute p-value**

Using t-distribution table with df $=11$, then $0.10 < p-value < 0.15$.

**Step 4: Conclusion**

Since p-value $> \alpha = 0.05$, we can't reject $H_0$. There is not
enough evidence to infer that the new fertilizer is better.
::::

## Two Sample Hypothesis Test on Proportions

Moreover, we can use hypothesis test to compare proportions from two
independent groups as well. We are going to start all these steps
again.\
**Step 1: Stating the Structure of Testing Hypothesis**

::: center
<figure>
<table>
<tbody>
<tr>
<td style="text-align: center;">Cases</td>
<td style="text-align: center;">Null Hypothesis</td>
<td style="text-align: center;">Alternative Hypothesis</td>
</tr>
<tr>
<td style="text-align: center;">1</td>
<td style="text-align: center;"><span
class="math inline"><em>p</em><sub>1</sub> − <em>p</em><sub>2</sub> = 0</span></td>
<td style="text-align: center;"><span
class="math inline"><em>p</em><sub>1</sub> − <em>p</em><sub>2</sub> &gt; 0</span></td>
</tr>
<tr>
<td style="text-align: center;">2</td>
<td style="text-align: center;"><span
class="math inline"><em>p</em><sub>1</sub> − <em>p</em><sub>2</sub> = 0</span></td>
<td style="text-align: center;"><span
class="math inline"><em>p</em><sub>1</sub> − <em>p</em><sub>2</sub> &lt; 0</span></td>
</tr>
<tr>
<td style="text-align: center;">3</td>
<td style="text-align: center;"><span
class="math inline"><em>p</em><sub>1</sub> − <em>p</em><sub>2</sub> = 0</span></td>
<td style="text-align: center;"><span
class="math inline"><em>p</em><sub>1</sub> − <em>p</em><sub>2</sub> ≠ 0</span></td>
</tr>
</tbody>
</table>
<figcaption>All possible cases of two sample hypothesis test on
proportions</figcaption>
</figure>
:::

In this case, we are interested in the difference of proportions.

**Step 2: Computing Test Statistics**

::: definition
To test the hypothesis $H_0 : p_1 = p_2$ first find the pooled
proportion $\hat{p}$ of successes in both samples combined. Then compute
the $Z*$ statistic:
$$z_* = \frac{\hat{p_1}-\hat{p_2}}{\sqrt{\hat{p}(1-\hat{p})(\frac{1}{n_1}+ \frac{1}{n_2})}} \sim N(0,1).$$
In this case, $\hat{p} = \frac{x_1 + x_2}{n_1+n_2}$, which is the number
of success with two groups combined. $n_1 \text{ and } n_2$ represent
sample size of each group respectively.
:::

**Step 3: Finding the P-value**

In terms of a variable $Z$ having the standard Normal distribution, the
approximate P-value for a test of $H_0$ against:

- $H_a: p_1 - p_2 > 0$ is $P(Z > Z_*)$;

- $H_a: p_1 - p_2 < 0$ is $P(Z < Z_*)$;

- $H_a: p_1 - p_2 \neq 0$ is $2 \cdot P(Z > |Z_*|)$.

**Step 4: Comparing P-value with $\alpha$-level**

If p-value is less than $\alpha$-level, then we reject the null
hypothesis ($H_0$) and accept the alternative hypothesis ($H_a$).
Otherwise, If p-value is greater than $\alpha$-level, then we do not
reject the null hypothesis ($H_0$) and reject the alternative hypothesis
($H_a$).\
**Step 5: Final Conclusion**

If we reject the null hypothesis, then we conclude that: there is
sufficient evidence to reject the null hypothesis. If we do not reject
the null hypothesis, then we conclude that: there is insufficient
evidence to reject the null hypothesis. Moreover:

- If $p_1 - p_2 > 0$: if we reject the null hypothesis under this
  alternative test, then $p_1 > p_2$.

- If $p_1- p_2 < 0$: if we reject the null hypothesis under this
  alternative test, then $p_1 < p_2$.

- If $p_1 - p_2 \neq 0$: if we reject the null hypothesis under this
  alternative test, then $p_1 \neq p_2$.

**Conditions of Two Sample Hypothesis Test on Proportions**

i\. Independent Response Assumption: Within each group , we need
independent responses from the cases. We cannot check that for certain,
but randomization provides evidence of independence. So, we need to
check the following:

- Randomization Condition: The data in each group should be drawn
  independently and at random from a population or generated by a
  completely randomized designed experiment.

- The 10 % Condition: If the data are sampled without replacement, the
  sample should not exceed 10 % of the population. If samples are bigger
  than 10 % of the target population, random draws are no longer
  approximately independent.

- Independent Groups Assumption: The two groups we are comparing must be
  independent from each other.

ii\. Sample Size Condition Each of the groups must be big enough. As
with individual proportions, we need larger group s to estimate
proportions that are near 0% and 100%. We check the success / failure
condition for each group.

- Success / Failure Condition: Both groups are big enough that at least
  10 successes and at least 10 failures have been observed in each group
  or will be expected in each (when testing hypothesis).

Note: Two-sided significance tests (later we will discuss this concept)
are robust against violations of this condition. In this case, we can
conduct significance tests with smaller sample sizes. In practice, the
two-sided significance test works well if there are at least five
successes and five failures in each sample.

::::: example
Nicotine patches are often used to help smokers quit. Does giving
medicine to fight depression help? A randomized double-blind experiment
assigned 244 smokers who wanted to stop to receive nicotine patches and
another 245 to receive both a patch and the anti-depression drug
bupropion. Results: After a y ear, 40 subjects in the nicotine patch
group had abstained from smoking, as had 87 in the patch-plus-drug
group. How significant is the evidence that the medicine increases the
success rate? State hypotheses, calculate a test statistic, use Table 6
to give its P-value, and state y our conclusion. (Use $\alpha = 0.01$)\
Solution:

**Step 1: State Hypothesis**

$H_0: p_1 = p_2$ and $H_a: p_1 < p_2$

**Step 2: Find test statistics**

$\hat{p_1} = \frac{40}{244} = 0.1639$ and
$\hat{p_2} = \frac{87}{245} = 0.3551$. Then
$\hat{p} = \frac{40+87}{244+245} = 0.2597$.

Now,
$z_* = \frac{\hat{p_1} - \hat{p_2}}{\sqrt{\hat{p}(1-\hat{p})(\frac{1}{n_1} + \frac{1}{n_2})}} = -4.82$

**Step 3: Compute p-value**

P-value $= P(Z < z_*) = P(Z < -4.82) < 0.0003$

**Step 4: Conclusion**

Since p-value $< 0.0003 < \alpha = 0.01$, we reject the null hypothesis
that $p_1 = p_2$. The data provide very strong evidence that bupropion
increases success rate.

**R-code:**

Input

::: tcolorbox
    successes=c(87, 40);

    totals=c(245, 244);

    prop.test(successes, totals, alternative="greater",correct=FALSE);
:::

Output

::: tcolorbox

    ## 
    ## 2-sample test for equality of proportions without 
    ## continuity correction 
    ## 
    ## data: x and n
    ## X-squared = 23.237, df = 1, p-value = 7.161e-07
    ## alternative hypothesis: greater 
    ## 95 percent confidence interval: 
    ## 0.1275385 1.0000000 
    ## sample estimates:
    ##    prop 1    prop 2 
    ## 0.3551020 0.1639344
:::
:::::

## Two Sample Hypothesis Test on Variances

Let's begin this type of hypothesis test with a case. The question is:
how do you know whether the homogeneity of variance assumption is
satisﬁed? One simple method involves just looking at two sample
variances. Logically, if two population variances are equal, then the
two sample variances should be very similar. When the two sample
variances are reasonably close, you can be reasonably conﬁdent that the
homogeneity assumption is satisﬁed and proceed with, for example,
Student t-interval. However, when one sample variance is three or four
times larger than the other, then there is reason for a concern. The
common statistical procedure for comparing population variances
$\sigma_1^2$ and $\sigma_2^2$ makes an inference about the ratio of
$\sigma_1^2$/$\sigma_2^2$.\
**Step 1: Stating the Structure of Testing Hypothesis**

::: center
<figure>
<table>
<tbody>
<tr>
<td style="text-align: center;">Cases</td>
<td style="text-align: center;">Null Hypothesis</td>
<td style="text-align: center;">Alternative Hypothesis</td>
</tr>
<tr>
<td style="text-align: center;">1</td>
<td style="text-align: center;"><span
class="math inline"><em>σ</em><sub>1</sub><sup>2</sup> = <em>σ</em><sub>2</sub><sup>2</sup></span></td>
<td style="text-align: center;"><span
class="math inline"><em>σ</em><sub>1</sub><sup>2</sup> &gt; <em>σ</em><sub>2</sub><sup>2</sup></span></td>
</tr>
<tr>
<td style="text-align: center;">2</td>
<td style="text-align: center;"><span
class="math inline"><em>σ</em><sub>1</sub><sup>2</sup> = <em>σ</em><sub>2</sub><sup>2</sup></span></td>
<td style="text-align: center;"><span
class="math inline"><em>σ</em><sub>1</sub><sup>2</sup> &lt; <em>σ</em><sub>2</sub><sup>2</sup></span></td>
</tr>
<tr>
<td style="text-align: center;">3</td>
<td style="text-align: center;"><span
class="math inline"><em>σ</em><sub>1</sub><sup>2</sup> = <em>σ</em><sub>2</sub><sup>2</sup></span></td>
<td style="text-align: center;"><span
class="math inline"><em>σ</em><sub>1</sub><sup>2</sup> ≠ <em>σ</em><sub>2</sub><sup>2</sup></span></td>
</tr>
</tbody>
</table>
<figcaption>All possible cases of two sample hypothesis test on
variances</figcaption>
</figure>
:::

**Step 2: Computing Test Statistics**

::: definition
The test statistics of two sample hypothesis test on variances is given
by: $$F_* = \frac{s_1^2}{s_2^2} \sim F_{n_1-1,n_2-1}.$$
:::

**Decision Rules**

- $H_0: \sigma_1^2 = \sigma_2^2$ and $H_a:\sigma_1^2 \neq \sigma_2^2$.
  If $F^* > F_{n_1-1,n_2-1,\alpha/2}$ or
  $F^* < F_{n_1-1,n_2-1,1 - \alpha/2}$, then we reject $H_0$. Otherwise,
  we do not reject it.

- $H_0: \sigma_1^2 = \sigma_2^2$ and $H_a: \sigma_1^2 > \sigma_2^2$. If
  $F^*> F_{n_1-1,n_2-1,\alpha}$ or $P(F_{n_1-1,n_2-1} > F^*)$ is too
  small, then we reject $H_0$. Otherwise, we do not reject it.

- $H_0: \sigma_1^2 = \sigma_2^2$ and $H_a: \sigma_1^2 < \sigma_2^2$. if
  $F^* < F_{n_1-1,n_2-1,1-\alpha}$ or $P(F_{n_1-1,n_2-1} < F^*)$ is too
  small, then we reject $H_0$. Otherwise, we do not reject it.
