# Law of Large Numbers

## Convergence in Probability

::: definition
The sequence of random variables $X_1, X_2, X_3, \ldots, X_n, \ldots$ is
said to **converge in probability** to the constant $c$, if for every
$\epsilon > 0$,
$$\lim_{n \to \infty} P\left( |X_n - c| \leq \epsilon \right) = 1$$ or
equivalently,
$$\lim_{n \to \infty} P\left( |X_n - c| > \epsilon \right) = 0$$

**Notation:** $X_n \xrightarrow{P} c$
:::

This concept plays a key role in the Law of Large Numbers, where the
sample mean of independent and identically distributed random variables
converges in probability to the population mean as the sample size
grows.

::: definition
Let $X$ be a random variable with finite mean $\mu$ and variance
$\sigma^2$. Then, for any $k > 0$,
$$P\left( |X - \mu| \geq k \right) \leq \frac{\sigma^2}{k^2}$$

*Using complements:*
$$P\left( |X - \mu| < k \right) \geq 1 - \frac{\sigma^2}{k^2}$$
:::

## Weak Law of Large Numbers (WLLN)

::: definition
Let $X_1, X_2, \ldots$ be a sequence of independent and identically
distributed random variables, each having finite mean $E(X_i) = \mu$ and
variance $\mathrm{Var}(X_i) = \sigma^2$. Then, for any $\epsilon > 0$,
$$P\left( \left| \frac{X_1 + X_2 + \cdots + X_n}{n} - \mu \right| \geq \epsilon \right) \to 0 \quad \text{as } n \to \infty$$

**Notation:** $\bar{X}_n \xrightarrow{P} \mu$
:::

### Proof of the Weak Law of Large Numbers (WLLN) {#proof-of-the-weak-law-of-large-numbers-wlln .unnumbered}

We aim to show that for every $\epsilon > 0$,
$$\lim_{n \to \infty} P\left( \left| \bar{X}_n - \mu \right| > \epsilon \right) = 0$$
where $\bar{X}_n$ is the sample mean of $n$ independent and identically
distributed (i.i.d.) random variables with
$$E(X_i) = \mu, \quad \text{and} \quad \mathrm{Var}(X_i) = \sigma^2.$$

Let $$\bar{X}_n = \frac{1}{n} \sum_{i=1}^{n} X_i.$$

By the Central Limit Theorem (CLT), we know that
$$\bar{X}_n \sim \mathcal{N} \left( \mu, \frac{\sigma^2}{n} \right).$$

Now, applying **Chebyshev's Inequality**, which states that for any
random variable $X$ with mean $\mu$ and variance $\sigma^2$,
$$P\left( |X - \mu| > k \right) \leq \frac{\sigma^2}{k^2} \quad \text{for } k > 0,$$
to $\bar{X}_n$, we set $k = \epsilon$, and obtain:
$$P\left( \left| \bar{X}_n - \mu \right| > \epsilon \right) \leq \frac{\mathrm{Var}(\bar{X}_n)}{\epsilon^2} = \frac{\sigma^2 / n}{\epsilon^2} = \frac{\sigma^2}{n \epsilon^2}.$$

Taking the limit as $n \to \infty$, we have:
$$\lim_{n \to \infty} P\left( \left| \bar{X}_n - \mu \right| > \epsilon \right) \leq \lim_{n \to \infty} \frac{\sigma^2}{n \epsilon^2} = 0.$$

Since probabilities are always non-negative, we conclude:
$$\lim_{n \to \infty} P\left( \left| \bar{X}_n - \mu \right| > \epsilon \right) = 0.$$

By the definition of convergence in probability,
$$\bar{X}_n \xrightarrow{P} \mu.$$ 0◻

::: example
Let $X_i$, for $i = 1, 2, 3, \ldots$, be independent Poisson random
variables with rate parameter $\lambda = 3$. Prove that:
$$\bar{X}_n \xrightarrow{P} 3$$

**Properties of Poisson Distribution:**
$$E(X_i) = \lambda, \quad \mathrm{Var}(X_i) = \lambda$$ In this case,
$\lambda = 3$, so: $$E(X_i) = \mathrm{Var}(X_i) = 3$$

**Proof:**

We know:
$$E\left( \frac{X_1 + X_2 + \cdots + X_n}{n} \right) = 3, \quad \text{and} \quad
\mathrm{Var}\left( \frac{X_1 + X_2 + \cdots + X_n}{n} \right) = \frac{3}{n}$$

Applying Chebyshev's Inequality:
$$P\left( \left| \frac{X_1 + X_2 + \cdots + X_n}{n} - 3 \right| \geq \epsilon \right) \leq \frac{3}{n \epsilon^2}$$

Taking the limit as $n \to \infty$:
$$P\left( \left| \frac{X_1 + X_2 + \cdots + X_n}{n} - 3 \right| \geq \epsilon \right) \to 0$$

**Conclusion:** $$\bar{X}_n \xrightarrow{P} 3$$
:::

```{r fig.cap="Simulation of running sample mean of Bernoulli \\(p = 0.5\\) trials over time", echo=FALSE, fig.align='center'}
set.seed(123)  # For reproducibility

n <- 10
p <- 0.5
sample <- rbinom(n, 1, p)
running_mean <- cumsum(sample) / (1:n)

# Plot the simulation
plot(1:n, running_mean, type = "o", col = "blue", pch = 1,
     xlab = "trial", ylab = "cumsum(sample)/trial",
     ylim = c(0, 1))
abline(h = p, lty = 2)
points(1:n, running_mean, col = "red", pch = 1, cex = 1.5)
```

**R Simulation Code (Single Sample Path):**

::: tcolorbox
    n = 10
    trial = seq(1, n, by = 1)
    sample = rbinom(n, 1, 1/2)

    plot(trial, cumsum(sample)/trial, type = "l", ylim = c(0,1), col = "blue")
    points(trial, cumsum(sample)/trial, col = "red")
    abline(h = 0.5, lty = 2, col = "black")
:::

```{r fig.cap="Simulation of 10 running sample means of Bernoulli \\(p = 0.5\\) trials converging over 100 trials", echo=FALSE, fig.align='center'}
set.seed(123)  # For reproducibility

n <- 100         # number of trials
p <- 0.5         # success probability
k <- 10          # number of simulations

colors <- rainbow(k)  # different colors for each line

# Plot setup
plot(1, type = "n", xlim = c(1, n), ylim = c(0, 1),
     xlab = "trial", ylab = "cumsum(sample1)/trial")

# Dashed line for true probability
abline(h = p, lty = 2)

# Run k simulations
for (i in 1:k) {
  sample <- rbinom(n, 1, p)
  running_mean <- cumsum(sample) / (1:n)
  lines(1:n, running_mean, col = colors[i])
}
```

**R Simulation Code (Multiple Sample Paths):**

::: tcolorbox
    n = 100
    trial = seq(1, 100, by = 1)

    sample1 = rbinom(n, 1, 1/2)
    sample2 = rbinom(n, 1, 1/2)
    sample3 = rbinom(n, 1, 1/2)
    sample4 = rbinom(n, 1, 1/2)
    sample5 = rbinom(n, 1, 1/2)
    sample6 = rbinom(n, 1, 1/2)
    sample7 = rbinom(n, 1, 1/2)
    sample8 = rbinom(n, 1, 1/2)

    colors = rainbow(8)


    plot(trial, cumsum(sample1)/trial, type = "l", col = colors[1], ylim = c(0,1))
    lines(trial, cumsum(sample2)/trial, col = colors[2])
    lines(trial, cumsum(sample3)/trial, col = colors[3])
    lines(trial, cumsum(sample4)/trial, col = colors[4])
    lines(trial, cumsum(sample5)/trial, col = colors[5])
    lines(trial, cumsum(sample6)/trial, col = colors[6])
    lines(trial, cumsum(sample7)/trial, col = colors[7])
    lines(trial, cumsum(sample8)/trial, col = colors[8])
    abline(h = 0.5, lty = 2, col = "black")
:::

### Empirical Probability Insight {#empirical-probability-insight .unnumbered}

The Law of Large Numbers gives us empirical probabilities. Consider
tossing a fair coin. Define the random variable $X$ as:

$$X = \begin{cases}
1 & \text{heads up} \\
0 & \text{tails up}
\end{cases}$$

Then as we sample more and more values of $X$, the sample mean
$\bar{X}_n$ converges in probability to $P(\text{heads up})$, that is:

$$\bar{X}_n \xrightarrow{P} P(\text{heads up})$$
