# Statistical Power


::: definition
The probability that a fixed level $\alpha$ significance test will
reject $H_0$ when a particular alternative value of the parameter is
true is called the **power** of the test against that alternative.
:::

The statistical power of a test is its ability to detect an effect if it
exists in reality.

It is the probability of correctly rejecting $H_0$ when $H_0$ is false
in reality.

$$\text{Power} = P(\text{reject } H_0 \mid H_0 \text{ false})$$

$$0 < \text{power} < 1$$

Power close to 1 (high power):\
Test is good at detecting effects.\
Power close to 0 (low power):\
Test is not reliable (i.e., we expect the test will not reject $H_0$
when $H_0$ is false).

Power is affected by:

- **The effect**\
  ([larger differences between reality and the null are easier to
  detect]{style="color: blue"})

- **Sample size**\
  ([larger samples increase power]{style="color: blue"})

- **Significance level** ($\alpha$)\
  ([as $\alpha$ increases, easier to reject $H_0$]{style="color: blue"})

- **Variability in data**\
  ([lower variability, higher power]{style="color: blue"})

### Type I and II Errors {#type-i-and-ii-errors .unnumbered}

It is possible to make an incorrect conclusion on a hypothesis test.\
**Type I:** Incorrectly reject $H_0$ when $H_0$ is true in reality.\
**Type II:** Incorrectly fail to reject $H_0$ when $H_0$ is false in
reality.\
**Reality vs Conclusion Table:**

::: center
+:-:+:----------------------------------------------------------------------------------------------------------------------:+
|   |                                               $H_0$ True                                  $H_0$ False                  |
|   |   -------------------------- -------------------------------------------- -------------------------------------------- |
|   |   **Reject $H_0$**             [Type I ($\alpha$)]{style="color: red"}     [No error ]{style="color: green!50!black"}  |
|   |   **Fail to reject $H_0$**    [No error ]{style="color: green!50!black"}    [Type II ($\beta$)]{style="color: red"}    |
+---+------------------------------------------------------------------------------------------------------------------------+
:::

***Note:*** Type I errors are generally considered worse.

Let $\beta$ be the probability of a Type II error. Then:

$$\text{Power} = 1 - \beta = 1 - P(\text{Type II})$$

::::: example
The cola maker determines that a sweetness loss is too large to accept
if the mean response for all tasters is $\mu = 1.1$. Will a 5%
significance test detect this?

**Hypotheses:** $$\begin{aligned}
H_0\!:&\ \mu = 0 \\
H_A\!:&\ \mu > 0
\end{aligned}$$

Assume: $$n = 10, \quad \sigma = 1, \quad \alpha = 0.05$$

**Step 1: Determine the rejection region.**

Since the test is one-sided with $\alpha = 0.05$, we find:
$$z_{\text{crit}} = 1.645 \quad \text{(from Z-table)}$$

We reject $H_0$ if: $$Z^* > 1.645$$

::: center

```{r fig.cap="Rejection region for $Z$ with $\\alpha = 0.05$", echo=FALSE, fig.align='center', out.width='60%'}
library(ggplot2)

# Data
x <- seq(-3, 3, length = 1000)
y <- dnorm(x)
x_reject <- seq(1.645, 3, length = 500)
y_reject <- dnorm(x_reject)

# Plot
ggplot(data.frame(x, y), aes(x, y)) +
  geom_line(color = "blue", linewidth = 1.2) +
  geom_area(data = data.frame(x = x_reject, y = y_reject), 
            aes(x = x, y = y), fill = "green", alpha = 0.4) +
  geom_vline(xintercept = 1.645, linetype = "dashed") +
  annotate("text", x = 1.645, y = 0.02, 
           label = expression(z[crit] == 1.645), hjust = -0.1, size = 3.5) +
  annotate("text", x = 0, y = 0.12, 
           label = expression("Do not reject " * H[0]), size = 4) +
  annotate("text", x = 2.3, y = 0.25, 
           label = expression("Reject " * H[0]), size = 4) +
  labs(x = "", y = "") +
  theme_minimal(base_size = 13)
```

:::

**Step 2: Find the equivalent critical value of** $\bar{x}$.

Since $\sigma$ is known,
$$Z = \frac{\bar{x} - \mu_0}{\sigma/\sqrt{n}} \quad \Rightarrow \quad
1.645 = \frac{\bar{x}_{\text{crit}} - 0}{1/\sqrt{10}} \Rightarrow \bar{x}_{\text{crit}} \approx 0.520$$

So we reject $H_0$ if $\bar{x} > 0.520$.

**Step 3: Calculate the power when $\mu = 1.1$ is true.**

$$P\left(Z > \frac{0.520 - 1.1}{1/\sqrt{10}}\right)
\approx P(Z > -1.83) = 1 - 0.0336 = 0.9664$$

**Interpretation:** There is a 96.6% chance the test correctly detects
$\mu = 1.1$.

::: center
```{r fig.cap="Power curve showing shaded rejection area under \\( H_A \\)", echo=FALSE, fig.align='center', out.width='60%'}
library(ggplot2)

# Parameters
mu_0 <- 0
mu_a <- 1.1
sigma <- 1 / sqrt(10)
x_vals <- seq(-1, 3, length.out = 1000)

# Distributions
df <- data.frame(
  x = x_vals,
  y0 = dnorm(x_vals, mean = mu_0, sd = sigma),
  yA = dnorm(x_vals, mean = mu_a, sd = sigma)
)

# Rejection region under H_A
x_reject <- seq(0.52, 3, length.out = 500)
y_reject <- dnorm(x_reject, mean = mu_a, sd = sigma)

# Critical value height (for vertical dashed line)
y_crit <- dnorm(0.52, mean = mu_a, sd = sigma)

ggplot(df, aes(x)) +
  geom_line(aes(y = y0), color = "blue", linewidth = 1.2) +
  geom_line(aes(y = yA), color = "red", linewidth = 1.2) +
  geom_area(data = data.frame(x = x_reject, y = y_reject),
            aes(x = x, y = y), fill = "green", alpha = 0.4) +
  geom_vline(xintercept = 0.52, linetype = "dashed") +
  annotate("text", x = 0.52, y = -0.01, label = expression(bar(x)[crit] == 0.520), vjust = 1.2, size = 3.5) +
  labs(x = "", y = "") +
  theme_minimal(base_size = 17) +
  annotate("text", x = 0.6, y = 0.8 * max(df$y0), label = expression(H[0] * ":  " * mu == 0), color = "blue", size = 4) +
  annotate("text", x = 1.6, y = 0.8 * max(df$yA), label = expression(H[A] * ":  " * mu == 1.1), color = "red", size = 4)
```

:::

This result can also be visualized using a power curve, which shows how
the probability of correctly rejecting $H_0$ increases with the true
mean $\mu$.

```{r fig.cap="Power curve for a one-sided test with points at \\( \\mu = 0.52 \\) and \\( \\mu = 1.1 \\)", echo=FALSE, fig.align='center'}
# Step 1: Define parameters
n <- 10
alpha <- 0.05
mu0 <- 0
sigma <- 1
z_crit <- qnorm(1 - alpha)  # = 1.645
xbar_crit <- mu0 + z_crit * sigma / sqrt(n)  # ≈ 0.520

# Step 2: Create a sequence of true means
mu_vals <- seq(0, 3, length.out = 100)

# Step 3: Calculate power for each mu
power <- 1 - pnorm((xbar_crit - mu_vals) / (sigma / sqrt(n)))

plot(mu_vals, power, type = "l", lty = 2, col = "blue", lwd = 2,
     ylab = "power", xlab = expression(mu), ylim = c(0, 1))

# Add red points at mu = 0.52 and mu = 1.1
points(c(0.52, 1.1), 1 - pnorm((xbar_crit - c(0.52, 1.1)) / (sigma / sqrt(n))),
       pch = 19, col = "red")
```
:::::

## Type I and Type II Errors

If we reject $H_0$ when in fact $H_0$ is true, this is a **Type I
error.**\
If we fail to reject $H_0$ when in fact $H_a$ is true, this is a **Type
II error.**\
The **significance level** $\alpha$ of any fixed level test is the
probability of a Type I error.\
The **power** of a test against any alternative is 1 minus the
probability of a Type II error for that alternative.\
The probability of making a Type II error is denoted by $\beta$.

::: tcolorbox
**Type I Error**\
$H_0$ is true, but sampling variation in the data leads you to reject
$H_0$, you've made a Type I error.\
When $H_0$ is true, a Type I error occurs if $H_0$ is rejected.\
**Type II Error**\
$H_0$ is false, but sampling variation in the data does not lead you to
reject $H_0$, you've made a Type II error.\
When $H_0$ is false, a Type II error occurs if $H_0$ is **NOT**
rejected.
:::

:::: example
According to Access and Support to Education and Training Survey (2008),
of 4,756 adult Canadians, 1,581 indicated that they worked at a job or
business at anytime (between July 2007 and June 2008), regardless of the
number of hours per week.

Is there evidence to suggest that the true proportion $p$ is greater
than 0.50?

$$\begin{aligned}
H_0\!:&\ p = 0.50 \\
H_a\!:&\ p > 0.50
\end{aligned}$$

**R Output**

::: tcolorbox
    prop.test(x = 1581, n = 4756, p = 0.50,
              alternative = "greater", correct = FALSE)

    ## 
    ##  1-sample proportions test without continuity
    ##  correction
    ## 
    ## data:  1581 out of 4756, null probability 0.5
    ## X-squared = 534.24, df = 1, p-value = 1
    ## alternative hypothesis: true p is greater than 0.5
    ## 95 percent confidence interval:
    ##  0.3212845 1.0000000
    ## sample estimates:
    ##         p 
    ## 0.3324222 
:::

**$P\text{-value} > \alpha = 0.05$**; we Fail to Reject $H_0$.

**This means we could be making a Type II error.**We indicated that
there is no evidence to conclude that the true proportion of adult
Canadians who worked at a job or business at anytime (between July 2007
and June 2008), regardless of the number of hours per week, was more
than 0.50 --- this conclusion implies that $H_0: p = 0.50$ is plausible,
but we could be wrong.
::::

::: example
Bottles of a popular cola are supposed to contain 300 milliliters (ml)
of cola. There is some variation from bottle to bottle because the
filling machinery is not perfectly precise. The distribution of contents
is Normal with standard deviation $\sigma = 3$ ml. Will inspecting 6
bottles discover underfilling?

The hypotheses are: $$\begin{aligned}
H_0\!:&\ \mu = 300 \\
H_a\!:&\ \mu < 300
\end{aligned}$$

A 5% significance test rejects $H_0$ if $z_* \leq -1.645$, where the
test statistic $z_*$ is: $$z_* = \frac{\bar{x} - 300}{3 / \sqrt{6}}$$

Power calculations help us see how large a shortfall in the bottle
contents the test can be expected to detect. Find the power of this test
against the alternative $\mu = 299$.

**Step 1. Write the rule for rejecting $H_0$ in terms of $\bar{x}$.**

We know that $\sigma = 3$, so the $z$ test rejects $H_0$ at the
$\alpha = 0.05$ level when:
$$z = \frac{\bar{x} - 300}{3 / \sqrt{6}} < -1.645$$

This is the same as:
$$\bar{x} < 300 - 1.645 \cdot \frac{3}{\sqrt{6}} \quad \Rightarrow \quad \bar{x} < 297.985$$

**Step 2. The power is the probability of this event under the condition
that the alternative $\mu = 299$ is true.**

To calculate this probability, standardize $\bar{x}$ using $\mu = 299$:
$$\begin{aligned}
\text{power} &= P(\bar{x} < 297.985 \mid \mu = 299) \\
&= P\left( Z < \frac{297.985 - 299}{3 / \sqrt{6}} \right) \\
&= P(Z < -0.83) = 0.2033
\end{aligned}$$
:::

## Using Power to Determine Sample Size

When designing a study, one of the most important decisions is how large
a sample to collect. If the sample size is too small, even meaningful
effects may go undetected due to low statistical power. On the other
hand, collecting an unnecessarily large sample can be inefficient and
costly. By using power calculations, researchers can determine the
minimum sample size needed to detect an effect of a given size with a
specified probability (power), while controlling for Type I error. This
section introduces how statistical power is used in planning and
justifying sample sizes before conducting a hypothesis test.

::: example
Suppose an experimenter wishes to test $$\begin{aligned}
H_0\!:&\ \mu = 100 \\
H_a\!:&\ \mu > 100
\end{aligned}$$ at the $\alpha = 0.05$ level of significance and wants
$1 - \beta$ to equal 0.60 when $\mu = 103$. What is the smallest (i.e.,
cheapest) sample size that will achieve that objective? Assume that the
variable being measured is Normally distributed with $\sigma = 14$.\
**Step 1. Write the rule for rejecting $H_0$ in terms of $\bar{x}_*$.**

By definition, $$\alpha = P(\text{we reject } H_0 \mid \mu = 100)
= P\left( \bar{X} > \bar{x}_* \mid \mu = 100 \right)
= P\left( Z > \frac{\bar{x}_* - 100}{14 / \sqrt{n}} \right) = 0.05$$

From the standard normal table, $P(Z > 1.645) = 0.05$, so:
$$\bar{x}_* = 100 + 1.645 \cdot \frac{14}{\sqrt{n}}$$

**Step 2. The power is the probability of this event under the condition
that the alternative $\mu = 103$ is true.**

To calculate this probability, standardize $\bar{x}$ using $\mu = 103$:

$$\begin{aligned}
\text{power} &= 1 - \beta = P(\bar{X} > \bar{x}_* \mid \mu = 103) \\
&= P\left( Z > \frac{\bar{x}_* - 103}{14 / \sqrt{n}} \right) = 0.60
\end{aligned}$$

From the standard normal table,
$P(Z > -0.25) \approx 0.5987 \approx 0.60$, so:
$$\frac{\bar{x}_* - 103}{14 / \sqrt{n}} = -0.25
\Rightarrow
\bar{x}_* = 103 - 0.25 \cdot \frac{14}{\sqrt{n}}$$

**Step 3. Solving for $n$**

From Steps 1 and 2:
$$100 + 1.645 \cdot \frac{14}{\sqrt{n}} = 103 - 0.25 \cdot \frac{14}{\sqrt{n}}$$

Solving for $n$:
$$\left( \frac{(1.645 + 0.25) \cdot 14}{103 - 100} \right)^2 \approx 78.2045$$

Therefore, a minimum of 79 observations must be taken to guarantee that
the hypothesis test will have the desired power of at least 0.60.
:::

::: example
A vending machine advertises that it dispenses 225 ml cups of coffee
($\sigma = 7$ ml). You believe the mean volume of coffee per cup is
something less than 225 ml. You plan to sample 40 cups of coffee from
this machine to test your hypothesis.

1.  If the true mean volume of coffee per cup is 223 ml, what is the
    power of your test at $\alpha = 0.05$? *(Homework)*

2.  How many coffee cups should you sample if you want to raise the
    power in part (a) to 0.80?

**Solution (b):**

**Step 1. Write the rule for rejecting $H_0$ in terms of $\bar{x}_*$.**

By definition: $$\begin{aligned}
\alpha &= P(\text{we reject } H_0 \mid \mu = 225) \\
&= P(\bar{X} < \bar{x}_* \mid \mu = 225) \\
&= P\left( Z < \frac{\bar{x}_* - 225}{7 / \sqrt{n}} \right) = 0.05
\end{aligned}$$

From the standard normal table, $P(Z < -1.645) = 0.05$, so:
$$\bar{x}_* = 225 - 1.645 \cdot \frac{7}{\sqrt{n}}$$

**Step 2. The power is the probability of this event under the condition
that the alternative $\mu = 223$ is true.**

To calculate this probability, standardize $\bar{x}$ using $\mu = 223$:

$$\begin{aligned}
\text{power} &= 1 - \beta = P(\bar{X} < \bar{x}_* \mid \mu = 223) \\
&= P\left( Z < \frac{\bar{x}_* - 223}{7 / \sqrt{n}} \right) = 0.80
\end{aligned}$$

From the table, $P(Z < 0.84) = 0.7995 \approx 0.80$, so:
$$\frac{\bar{x}_* - 223}{7 / \sqrt{n}} = 0.84
\quad \Rightarrow \quad
\bar{x}_* = 223 + 0.84 \cdot \frac{7}{\sqrt{n}}$$

**Step 3. Solving for $n$**

From Steps 1 and 2:
$$225 - 1.645 \cdot \frac{7}{\sqrt{n}} = 223 + 0.84 \cdot \frac{7}{\sqrt{n}}$$

Solving for $n$:
$$n = \left( \frac{(1.645 + 0.84) \cdot 7}{225 - 223} \right)^2 \approx 75.6465$$

Therefore, a minimum of 76 observations must be taken to guarantee that
the hypothesis test will have the desired precision.
:::

::: example
A newsletter reports that 90% of adults drink milk. The researchers are
interested in investigating if fewer than 90% of adults drink milk (at
$\alpha = 0.05$). They collect a random sample of 200 adults in a
certain region.

1.  Calculate power of the test if the percentage of adults who drink
    milk is really 85%.

    $$\begin{aligned}
    \text{Under } H_0 &: \hat{p}^* = 0.90 - 1.645 \sqrt{\frac{0.90(1 - 0.90)}{200}} = 0.8651 \\
    \text{Power} &= P(\text{Reject } H_0 \mid p = 0.85) \\
    &= P(\hat{p} < 0.8651 \mid p = 0.85) \\
    &= P\left(Z < \frac{0.8651 - 0.85}{\sqrt{\frac{0.85(1 - 0.85)}{200}}} \right) \\
    &= P(Z < 0.5980) \approx 0.7250
    \end{aligned}$$

    **R Output**

    ::: tcolorbox
        pnorm(0.5980, mean = 0, sd = 1)
        # [1] 0.72508
    :::

2.  Calculate beta if the percentage of adults who drink milk is really
    85%.

    $$\begin{aligned}
    \beta &= 1 - \text{Power} = 1 - 0.725 = 0.275
    \end{aligned}$$

3.  How many adults should you sample if you want to raise the power in
    part (a) to 0.80?

    **Step 1: Determine rejection cutoff under** $H_0 : p = 0.90$
    $$\begin{aligned}
    \hat{p}^* = 0.90 - 1.645 \sqrt{\frac{0.90(1 - 0.90)}{n}}
    \end{aligned}$$

    **Step 2: Set power to 0.80 under** $p = 0.85$ $$\begin{aligned}
    P\left( \frac{\hat{p} - 0.85}{\sqrt{\frac{0.85(1 - 0.85)}{n}}} < \frac{\hat{p}^* - 0.85}{\sqrt{\frac{0.85(1 - 0.85)}{n}}} \right) &= 0.80 \\
    \frac{\hat{p}^* - 0.85}{\sqrt{\frac{0.85(1 - 0.85)}{n}}} &= 0.8416 \\
    \hat{p}^* &= 0.85 + 0.8416 \sqrt{\frac{0.85(1 - 0.85)}{n}}
    \end{aligned}$$

    **R Output**

    ::: tcolorbox
        qnorm(0.80, mean = 0, sd = 1)
        # [1] 0.8416212
    :::

    **Step 3: Equating expressions for** $\hat{p}^*$ $$\begin{aligned}
    0.90 - 1.645 \sqrt{\frac{0.90(1 - 0.90)}{n}} &= 0.85 + 0.8416 \sqrt{\frac{0.85(1 - 0.85)}{n}} \\
    0.05 &= \left(1.645 \sqrt{0.90(0.10)} + 0.8416 \sqrt{0.85(0.15)}\right) \cdot \frac{1}{\sqrt{n}} \\
    \sqrt{n} &= \frac{0.4935 + 0.3005}{0.05} = 15.87 \Rightarrow n \approx 253
    \end{aligned}$$
:::

:::: example
A newsletter reports that 90% of adults drink milk. The researchers are
interested in investigating if less than 90% of adults drink milk (at
$\alpha = 0.05$). They collect a **random sample of 100 adults** in a
certain region.\
Calculate power of the test if the percentage of adults who drink milk
is really 85%.

We test: $$\begin{aligned}
H_0 &: p = 0.90 \\
H_a &: p < 0.90
\end{aligned}$$

The rejection region is determined by: $$\begin{aligned}
\alpha &= 0.05 = P(\text{reject } H_0 \mid H_0 \text{ is true}) \\
P(Z < -1.645) &= 0.05 \quad \text{(Z critical value is -1.645)}
\end{aligned}$$

Now calculate power under the alternative $p = 0.85$:

$$\begin{aligned}
\text{Power} &= P(\text{reject } H_0 \mid H_0 \text{ is false}) \\
&= P\left( \frac{\hat{p} - 0.90}{\sqrt{\frac{0.90(1 - 0.90)}{100}}} < -1.645 \,\middle|\, p = 0.85 \right) \\
&= P(\hat{p} < 0.85065 \mid p = 0.85) \\
&= P\left( Z < \frac{0.85065 - 0.85}{\sqrt{\frac{0.85(1 - 0.85)}{100}}} \right) \\
&= P(Z < 0.0182) \approx 0.5072
\end{aligned}$$

**R Output**

::: tcolorbox
    pnorm(0.0182, mean = 0, sd = 1)
    # [1] 0.5072603
:::
::::

:::: example
A newsletter reports that 90% of adults drink milk. The researchers are
interested in investigating if less than 90% of adults drink milk (at
$\alpha = 0.05$). They collect a **random sample of 50 adults** in a
certain region.\
Calculate power of the test if the percentage of adults who drink milk
is really 85%.

$$\begin{aligned}
H_0 &: p = 0.90 \\
H_a &: p < 0.90
\end{aligned}$$

The rejection region is determined by: $$\begin{aligned}
\alpha &= 0.05 = P(\text{reject } H_0 \mid H_0 \text{ is true}) \\
P(Z < -1.645) &= 0.05 \quad \text{(Z critical value is -1.645)}
\end{aligned}$$

**Power:** $$\begin{aligned}
\text{Power} &= P(\text{reject } H_0 \mid H_0 \text{ is false}) \\
&= P\left( \frac{\hat{p} - 0.90}{\sqrt{\frac{0.90(1-0.90)}{50}}} < -1.645 \;\middle|\; p = 0.85 \right) \\
&= P(\hat{p} < 0.85065 \mid p = 0.85) \\
&= P\left( Z < \frac{0.85065 - 0.85}{\sqrt{\frac{0.85(1 - 0.85)}{50}}} \right) \\
&= P(Z < -0.3921) \\
&\approx 0.3475
\end{aligned}$$

**R Output**

::: tcolorbox
    pnorm(-0.3921, mean = 0, sd = 1)
    # [1] 0.3474922
:::
::::

If we keep $\alpha$ at the same size, larger sample sizes increase the
power of the test because sampling variability (sampling distributions)
are much narrower.

::: tcolorbox
**If $H_a : \mu > \mu_0$, then**
$$\text{Power}(\mu_a) = \pi(\mu_a) = P \left[ Z \geq Z_{\alpha} + \frac{\sqrt{n}(\mu_0 - \mu_a)}{\sigma} \right]$$

**where** $\mu_a > \mu_0$.

**If $H_a : \mu < \mu_0$, then**
$$\text{Power}(\mu_a) = \pi(\mu_a) = P \left[ Z \leq -Z_{\alpha} + \frac{\sqrt{n}(\mu_0 - \mu_a)}{\sigma} \right]$$

**where** $\mu_a < \mu_0$.

**If $H_a : \mu \ne \mu_0$, then** $$\begin{aligned}
\text{Power}(\mu_a) &= \pi(\mu_a) \\
&= 1 - P \left[ -Z_{\alpha/2} + \frac{\sqrt{n}(\mu_0 - \mu_a)}{\sigma} \leq Z \leq Z_{\alpha/2} + \frac{\sqrt{n}(\mu_0 - \mu_a)}{\sigma} \right]
\end{aligned}$$

**where** $\mu_a \ne \mu_0$.

**You need to know:**

- Standard deviation, $\sigma$

- Significance level, $\alpha$

- Effect size to detect, $\mu_0 - \mu_a$
:::

Calculations of power (or of error probabilities) are useful for
planning studies because we can make these calculations before we have
any data. Once we actually have data, it is more common to report a
P-value rather than a reject-or-not decision at a fixed significance
level $\alpha$. The P-value measures the strength of the evidence
provided by the data against $H_0$. It leaves any action or decision
based on that evidence up to each individual. Different people may
require different strengths of evidence.
