\chapter{Statistical Power}

\begin{definition}[Power of a Test]
The probability that a fixed level $\alpha$ significance test will reject $H_0$ when a particular alternative value of the parameter is true is called the \textbf{power} of the test against that alternative.
\end{definition}

The statistical power of a test is its ability to detect an effect if it exists in reality.

It is the probability of correctly rejecting $H_0$ when $H_0$ is false in reality.

\begin{equation*}
\text{Power} = P(\text{reject } H_0 \mid H_0 \text{ false})
\end{equation*}

\begin{equation*}
0 < \text{power} < 1
\end{equation*}

Power close to 1 (high power):\\
\quad Test is good at detecting effects.\\

Power close to 0 (low power):\\
\quad Test is not reliable (i.e., we expect the test will not reject $H_0$ when $H_0$ is false).

\vspace{1em}
Power is affected by:

\begin{itemize}
 \item \textbf{The effect} \\ (\textcolor{blue}{larger differences between reality and the null are easier to detect})

  \item \textbf{Sample size} \\ (\textcolor{blue}{larger samples increase power})

  \item \textbf{Significance level} ($\alpha$) \\ (\textcolor{blue}{as $\alpha$ increases, easier to reject $H_0$})

  \item \textbf{Variability in data} \\ (\textcolor{blue}{lower variability, higher power})
\end{itemize}
\section*{Type I and II Errors}

It is possible to make an incorrect conclusion on a hypothesis test. \\

\textbf{Type I:} Incorrectly reject $H_0$ when $H_0$ is true in reality. \\
\textbf{Type II:} Incorrectly fail to reject $H_0$ when $H_0$ is false in reality. \\


\noindent \textbf{Reality vs Conclusion Table:}
\vspace{1em}
\begin{center}
\small
\renewcommand{\arraystretch}{1.8}
\setlength{\tabcolsep}{1.5em}

\text{Reality}

\vspace{0.5em}

\begin{tabular}{c c}
\hspace*{2em}\rotatebox[origin=c]{90}{\textbf{Conclusion}} &
\begin{tabular}{|l|c|c|}
\hline
& $H_0$ True & $H_0$ False \\
\hline
\textbf{Reject $H_0$} & \textcolor{red}{Type I ($\alpha$)} & \textcolor{green!50!black}{No error \checkmark} \\
\hline
\textbf{Fail to reject $H_0$} & \textcolor{green!50!black}{No error \checkmark} & \textcolor{red}{Type II ($\beta$)} \\
\hline
\end{tabular}
\end{tabular}
\end{center}



\vspace{1em}
\textbf{\textit{Note:}} Type I errors are generally considered worse.

\vspace{0.5em}
\noindent Let $\beta$ be the probability of a Type II error. Then:

\[
\text{Power} = 1 - \beta = 1 - P(\text{Type II})
\]
