\subsection{Exercises}
\begin{enumerate}
\item For each of the following refection regions,sketch the sampling distribution for $z$ and indicate the location of the rejection region.

\begin{enumerate}
\item $z>1.96$
\item $z>2.575$
\item $z<-1.28$
\item $z<-1.645$ or $z>1.645$
\item In part a-d,what is the probability that a Type I error will be made?
\end{enumerate}
\item Suppose you are interested in conducting the statistical test of $H_0:\mu=225$ against $H_a:\mu>225$,and you have decided to use the following decision rule:Refect $H_0$ if the sample mean of a random sample of 81 items is more than 270.Assume that the standard deviation of the population is 63.
\begin{enumerate}
\item  Express the decision rule in terms of $z$.
\item  Find $\alpha$,the probability of making a Type I error,by using this decision rule.
\end{enumerate}
\item A random sample of 100 observations from a population with standard deviation 60 yielded a sample mean of 110.
  \begin{enumerate}
\item Test the null hypothesis that $\mu=100$ against the alternative hypothesis that $\mu>100$ using $\alpha=0.05$.Interpret the results of the test.
\item Test the null hypothesis that $\mu=100$ against the alternative hypothesis that $\mu\neq100$ using $\alpha=0.05$.Interpret the results of the test.
\item Compare the results of the two tests you conducted.Explain why the results differ.
\end{enumerate}  
\item A random sampel of 64 observations produced the following summary statistics:$\bar{x}=0.323$ and $s^2=0.034$.
\begin{enumerate}
\item Test the null hypothesis that $\mu=0.36$ against the alternative hypothesis that $\mu<0.36$ using $\alpha=0.10$
\item Test the null hypothesis that $\mu=0.36$ against the alternative hypothesis that $\mu\neq0.36$ using $\alpha=0.10$.Interpret the results of the test.
\end{enumerate} 
\end{enumerate}
